A Pilot Multicenter Cluster Randomized Trial to Compare The Effect of Trauma Life Support Training Programs on Patient and Provider Outcomes

### 2022-12-02 pilot-publication/manuscript.Rmd
Finalise early but incomplete draft

## Corresponding author
Martin Gerdin Wärnberg (MGW)

@@

@@

## Word Count
<!-- 4494 -->

## Role of study sponsor and funders
The funding sources had no role in the design of this study and will not
have any role during its execution, analyses, interpretation of the
data, or decision to submit results.

@@

**Introduction** Several trauma life support programs aim to improve trauma outcomes but there is no evidence from controlled trials to show the effect of these programs on patient outcomes. We conducted a pilot study to assess the feasibility of conducting a cluster randomised controlled trial comparing the effect of Advanced Trauma Life Support (ATLS) and Primary Trauma Care (PTC) with standard care on patient outcomes. **Methods and analysis** We piloted a pragmatic three-armed parallel, cluster randomised, controlled trial in India, where neither of these programs are routinely taught. We recruited tertiary hospitals and included trauma patients and residents managing these patients. Two hospitals were randomised to ATLS, two to PTC, and three to standard care. The primary outcome was all cause mortality at 30 days from the time of arrival to the emergency department. We conducted community consultations in parallel with the pilot trial. **Ethics and dissemination** We obtained ethical approval from all participating centres. **Results** Between `r start.date` and `r end.date` we included `r n.patients` patients and `r n.residents` residents from `r n.centres` centres. The patient recruitment rate was X and the resident recruitment rate was X. The loss to follow up rate was X. The all-cause 30 day mortality was X in the ATLS arm, X in the PTC arm, and X in the standard care arm. **Conclusions** Conducting a full-scale cluster randomised controlled trial comparing the effects of ATLS, PTC, and standard care on patient outcomes should be feasible after incorporating key lessons from this pilot.

@@

# Article Summary

@@

- Cluster randomized controlled trial comparing the effect of ATLS, PTC and standard care on patient and provider outcomes.
- Prospective data collection with direct observations by dedicated project officers.
- Participating centers' heterogeneity may affect the study estimates and bias the results.

@@

We will perform a pilot study that aims to assess the feasibility of
conducting a cluster randomised controlled trial comparing ATLS and
PTC with standard care. Recent methodological guidelines indicate that
the design of efficient cluster randomised controlled trials requires
data on probable or target effect sizes, proportion of participants
with the outcome (if binary), and the intracluster correlation
coefficient [@Hemming2017]. The objectives of this pilot study will be
to:

@@

## Trial Design
This study will pilot a pragmatic three-armed parallel, cluster
randomised, controlled trial. There will be two intervention arms,
ATLS and PTC training, and one control arm, standard care. We will
collect data for four months in all three arms, first during a one
month observation phase and then during a three month intervention
phase (or continued observation in the control arm). This design will
allow us to assess outcomes both as final values and as change from
baseline. Our study is a pilot study because its objectives involves
estimating quantities, such as the probable effect sizes, proportion
of participants with the outcome (if binary), and the intracluster
correlation coefficient, needed for the sample size calculations of a
full-scale trial [@Hemming2017]. The full-scale trial will be planned
regardless of the effect-sizes identified in this pilot
study. This pilot study will also establish how many
participants that can be enrolled, as well as likely drop out rates,
and the feasibility of collecting primary and secondary outcomes.

## Study Setting
We conducted this pilot in seven Indian tertiary hospitals, where
neither ATLS, PTC, nor any other trauma life support training program
is routinely taught. India is the world's second most populous country
and has 20% of the world's trauma deaths. The trauma system is still
developing, with limited prehospital care, and the in hospital trauma
mortality as well as the proportion of preventable deaths remain
high. Lack of standard trauma training for healthcare providers,
limited hospital resources, inadequate processes of care, overcrowding
emergency departments - are some of the factors that contribute to the
high mortality and morbidity. During recent years efforts have been
made to improve hospital trauma care, through capacity building for
trained trauma care providers, augmenting facilities, and developing
care protocols within the hospitals. We will conduct this trial in
India because training providers in a trauma life support program is
not yet the standard.

The seven hospitals were King Edward Memorial Seth G. S. Medical
College and K.E.M. Hospital, Mumbai; Lokmanya Tilak Municipal Medical
College and General Hospital, Mumbai; HBT Medical College And Dr. R N
Cooper Municipal General Hospital, Mumbai; Medica Superspecialty
Hospital, Kolkata; Medical College Kolkata, Kolkata; Sir Nil Ratan
Sircar Medical College & Hospital, Kolkata; Postgraduate Institute of
Medical Education & Research, Chandigarh.

## Eligibility Criteria for Participants and Clusters
There will be two groups of participants: patients and resident
doctors.

### Patient Participants
Adults (15 years or older) who present to the emergency department at
participating hospitals with a history of trauma. History of trauma is
here defined as having any of the external causes of morbidity and
mortality listed in block V01-Y36, chapter XX of the International
Classification of Disease version 10 (ICD-10) codebook as reason for
presenting. We will explore intervention effects across the following
clinical subgroups: men, women, blunt multisystem trauma, penetrating
trauma, shock, severe traumatic brain injury, and elderly, as defined by
Hornor et al [@Hornor2018]. The consent form for patients are available
as Supplemental Material 1.

### Resident Doctor Participants
Resident doctors doing their speciality training in surgery (in India
it is very rare to have emergency medicine managing trauma patients in
the emergency department, why we decide to focus on surgical units in
this pilot), who manage trauma patients in the emergency department,
and who are expected to remain in the participating hospitals for at
least one year. To facilitate administration each surgical department
is divided into units, which manages the out patient department,
emergency department, operating rooms etc on different days each
week. One or two, out of typically six, units'
residents will be selected from each hospital. One unit consists of at
least three faculty and three to twelve residents.

To be eligible, units should have a maximum of 25% of the doctors
trained in either ATLS, PTC, or similar training programs before the
start of the pilot (hospitals that have so far agreed to participate
have no or single current residents trained in any program). Those
residents who have received training in the last five years will be
considered as trained. The figure of 25% was decided through consensus
in the research team, to balance feasibility and contamination of
results. We will select the units by conducting a prior survey to
ascertain this criteria. Consent will be sought from the residents in
each of the intervention groups before they undergo the ATLS or PTC
training. The consent form for residents are available as
Supplemental Material 2. We will not ask
for consent from residents at the units in the control hospitals as
their practice will not be affected by this pilot and we will not
collect any personal identifiable data on them. This is in line with
ethical regulations in the study setting.

### Clusters
Indian tertiary care hospitals that admit 400-800 adult patients with
trauma each year. We randomise on the cluster (hospital) level to avoid
contamination between intervention and control arms. To be eligible for
inclusion hospitals have to provide the following services round the
clock: operation theatres, X-ray, CT, and ultrasound facilities, and
blood bank. In addition the baseline admission rate should be more than
35 adult patients with major trauma per month.

## Interventions
In each intervention arm one or two units', out of typically
six, residents per hospital providing
emergency care to trauma patients will be trained in either ATLS or
PTC. For the purpose of this pilot study, we will target to train a
minimum of 75% of residents in each unit. If residents drop out or
change units after training but before data collection is completed we
will conduct additional training if needed to meet the 75% criterion.
We will not train the units' faculty, as they are typically not
involved in the initial management of trauma patients.

The ATLS training will be conducted in the nearest ATLS certified
training centre in India according to the standard ATLS curriculum
[@acsAtls2018]. The PTC training will be arranged in hospitals
randomized to the PTC arm, according to the standard PTC curriculum
[@ptcfPtc2018]. These courses will be conducted over a period of
2.5 to 3 days. The residents certified "pass" will be considered as
trained in respective courses.

@@

### Modifications
Both ATLS and PTC are standard training programs with fixed curricula
[@acsAtls2018; @ptcfPtc2018]. We will not modify the
delivery or content of these programs during this pilot.

### Adherence
The intervention is the training in either ATLS or PTC. Participants are
required to adhere to, i.e. participate in, the training, to be eligible
for passing. We will not consider adherence to training contents during
care delivery as adherence to the trial intervention, but rather as a
provider level outcome.

@@

#### Baseline Training
The care provided by all participating hospitals at baseline is based on
the training curriculum formulated by The National Medical Council of
India for post graduation in General Surgery
[@mciMsSurgery2019]. Regarding trauma, these guidelines state
that the student should:

@@

#### Standard of Care
At most medical colleges in India trauma patients present to the
emergency department where they are assessed by a doctor and referred
to the surgical bay for further management. In the surgical bay a
second or third year general surgery resident sees all the major
trauma and provide the initial care, including
initiating treatment and investigations.  This resident informs the
consultant on call who is generally an Assistant Professor. Most
procedures like intercostal drainage, open wound suturing, intubation
etc. would be done in the surgical resuscitation area, by the surgical
resident.

@@

## Outcomes
Our pilot study include both participant and feasibility
outcomes. Prior to deciding on these participant outcomes we searched
the Core Outcome Measures in Effectiveness Trials (COMET) Initiative's
database but were unable to identify appropriate core outcome sets for
our populations of participants.

The primary participant outcome will be all cause mortality within 30
days from the time of arrival to the emergency department. The
primary outcome and most secondary outcome will be assessed and
compared both as final values and as change from
baseline. All outcomes that pertain to the
individual participant level are detailed in Supplemental Material
3. We decided to include a large number of outcomes,
including some more exploratory, so that we can test their feasibility
and relevance. We may remove secondary participant outcomes during the
course of the pilot study, if they prove to be too difficult to
collect. If we remove outcomes we will document the reasons for doing
so.

We will also assess the following feasibility outcomes, which pertain
both to overall study population as well as to the individual cluster
level:

- Recruitment rate. For both patients and residents this will equal
  the proportion of participants enrolled, out of the total number of
  eligible participants, over the course of the pilot study.
- Lost to follow up rate. This will apply only to patients and equal
  the proportion of patients that do not complete 30 day follow up,
  out of all enrolled patients, over the course of the pilot study.
- Pass rate. This will apply only to residents in the intervention
  arms and equal the proportion of residents that pass the training
  programme, out of the total number of trained residents, over the
  course of the pilot study.
- Missing data rate. This will apply to each outcome and variable and
  equal the proportion of missing data, over the course of the pilot
  study.
- Differences in distributions of observed and extracted data. This
  will apply to each outcome and variable and will compare the
  distributions of data collected by observations versus extracted
  from hospital records. For quantitative variables this will be the
  difference in means, standard deviations, medians, interquartile
  ranges, and ranges. For qualitative variables this will be the
  differences in absolute counts and percentages, across categories.

@@

### Patients
Patients will be screened for eligibility as they arrive at the
emergency department. Eligible patients will be approached in the
emergency department to consent to follow up, if they are
conscious. If they are unconscious a patient representative will be
approached to consent to follow up. Once the patient is conscious we
will approach the patient to affirm the patient representative's
consent. We will follow up patients at discharge, at 24 hours after
arrival at the emergency department, and at 30 days after arrival at
the emergency department.

### Residents
Surgical units will be screened for eligibility once hospitals confirm
their participation. All residents in eligible units will be approached
to consent to training if their hospital is randomised to either of the
intervention arms. Training will be conducted as soon as possible after
the study starts. Resident participants will be followed up 30 days
after training, if they are in the intervention arms, or 30 days after
the study started, if they are in the control arm.

## Sample size
Given budget and time constraints, including the rotation of surgical
units in Indian hospitals (which often happen on a six months basis)
the feasible data collection period is four months. Each of the
surgical units see 2-4 trauma patients per week. If we select a
minimum of one unit per hospital then each hospital will enrol 8-16
patients per month and 32-64 patients during the four months of this
pilot. With a 20% attrition rate we expect each hospital to enrol
26-51 patients, coming to a total sample size of between 156 and 306
patients for this pilot study.

## Recruitment
To ensure adequate recruitment we only approach hospitals with trauma
volumes high enough to allow us to reach the sample size goals detailed
above. Patients will be enrolled by a dedicated project officer as they
arrive at the emergency department. The recruitment period will be three
months. Recruitment will be monitored weekly through online conferences.
No financial or non-financial incentives will be provided to trial
investigators or participants for enrolment.

@@

### Implementation
The random allocation sequence was generated by MGW. Clusters were
enrolled by the project's core team, who also enrolled
clusters. Patient participants will be included if they present during
the project officers shift. Resident participants are enrolled if they
are in the units selected for training. We will use simple random
sampling to select units if there are more than two eligible units in
a hospital. For patient participants consent for follow up is sought
after randomisation from patients or patient relatives as
appropriate. For resident participants consent is sought before
randomisation. If residents in a unit decline to participate, so that
the target of training 75% of residents in a given unit cannot be met,
another unit will be selected for participation.

## Blinding
It will not be possible to blind investigators or participants to
interventions. We will not blind the data analysts during this pilot,
but we plan to blind the data analysts during the full-scale
trial.

## Data Collection
Data collection will start one month before the training is delivered,
to establish a baseline. A variability of three months of the date when
data collection is started between hospitals will be accepted. Each
participating hospital will have a dedicated project officer to collect
data. The project officers will have a masters in a health science field
and should have experience in data collection.

Because participating residents are assigned designated days for trauma
care for a period of 6 months, data will be collected during those
particular days and shifts when these trained doctors are in the
emergency department. The project officers will collect data both by
observing the care delivered and by interviewing the participants, and
by extracting data from hospital records.

Data collection will continue for a minimum of three months after
training. The research officers will collect data of all patients, who
present with trauma in the surgical bay during their duty hours. Those
patients who are admitted will be followed up for complications and
other in-hospital outcome measures, for example length of stay. Patients
who are not admitted will be followed up telephonically for mortality
outcomes and quality of life outcomes. The follow up period will be 30
days. The project officers will make at most three attempts to reach a
participant or participant representative telephonically, after which
the data will be recorded as missing.

The project officer will administer the study information and informed
consent (consent will only be sought for data collection including
follow up) to the patient, or the patient's representative as
appropriate, once the patient is stabilised. They will continue to
collect data once they have received the consent.

Details of data of those patients/relatives not willing to give consent
will be removed from the analysis. The number of patients who opt out
from data collection will be collected, as well as limited data on their
age and sex. Patients will be followed up in the ward regularly for the
various outcome variables. They will also be followed up telephonically
after they have been discharged.

## Variables
The project officers will collect data on demographics, time of injury
to arrival at the participating hospital, time to recording vital
signs, vital signs, and times to and management details including
imaging and surgery. Details of any injury sustained will be collected
and coded using ICD 10 and the Abbreviated Injury Scale (AIS). For ICD
10 coders will undergo the WHO online ICD 10 training module and for
AIS they will be accredited. Based on AIS we will calculate the Injury
Severity Score (ISS) and the New ISS (NISS). Supplemental Material 4
contains a full variable list, with definitions.

## Patient and public involvement
In this study, we will conduct community consultations to collect
inputs from patients, their caregivers, patient groups, and resident
doctors to be used in the selection of outcome measures and
implementation of the full-scale trial; following the Guidance for
Reporting Involvement of Patients and the Public (GRIIPS 2)
[@Staniszewska2017].

During the pilot study, interviews will be conducted with
post-discharge trauma patients and their caregivers to identify
outcomes most relevant to them. These patients will be identified
through the medical registers of the participating hospitals,
contacted through telephone, and after receiving their consent be
interviewed as per their convenience. Their consent form is available
as Supplemental Material 5.  Additionally, members from non-government
organizations working with trauma patients and the hospital Social
Service Section will also be contacted for their views on contextual
patient-centred outcomes for trauma patients. Their consent form is
available as Supplemental Material 6. For feasibility, these
interviews will be held in each of the cities where the participating
centres are located. The commonest patient-centred outcomes reported
across all the locations will be incorporated into the evaluation of
the effects of the different training programs and standardized care
on patient outcomes.

Similarly, the inputs of resident doctor participants at each
participating centre will be collected during the pilot study. A
discussion and periodic surveys will be conducted to document any
challenges or suggestions they may have in the scheduling or
implementation of the training programs. These inputs will be
incorporated in the final study.

A summary of the findings of the study as well as their inputs will be
shared with those who participated in the interviews and surveys. A
meeting will be held with the patient participants, at each city, where
the changes in the measured patient-centered outcomes would be presented
to them. Another meeting will be held with the resident doctors at each
hospital to present the confidence of the residents after being trained.
Any suggestions and reflections from the participants during the
meetings will be used as inputs for planning the final study.

## Data management
We will supply an online data collection tool, accessible only over a
virtual private network (VPN), for each participating hospital to upload
pseudonymised data to secure servers. Data validation techniques like
restricted values or values of a specific range will be used to avoid
ambiguous data entries and ensure the validity of the data. Ambiguous
responses, data errors, if any, will be resolved after discussion with
the core team during weekly meetings. An instruction manual or codebook
for data variables will be prepared to ensure consistency in data entry.
This manual will be referred to during the project data collection and
variable descriptions are visible for each variable in the online data
collection tool. Pseudonymised data will be stored at the centralised
server. The data will be accessible by the project's principal
investigator or by delegation of the project principal investigator
only.

## Data monitoring
Weekly meetings with the core team and project officers will take
place and for this meeting a data status report will be automatically
generated highlighting missing data and number of patients awaiting
followup. We conducted an interim analysis on October 12, 2022, and
decided to complete the study as participants were consenting to be
included in the study and key variables including mortality outcomes
could be collected. No outcomes were dropped. We did not use a data
monitoring committee in the pilot study due to its limited scope.

## Statistical Methods
We will analyse all pilot data using descriptive
statistics. Quantitative variables will be summarised as mean +/-
standard deviation, median, interquartile range and range. Qualitative
variables will be presented as absolute numbers and
percentages. Feasibility outcomes will be summarised both on the
overall sample level as well as on the individual cluster level. We
will use an empty generalised linear mixed model<!-- see [@Sommet2017]
--> to estimate the intracluster correlation coefficient.

We will compare participant outcomes in three combinations of trial
arms: ATLS versus PTC, ATLS versus standard care, and PTC versus
standard care. In each combination we will compare both differences in
final values and differences in change from baseline. For example, for
the primary participant outcome of all cause mortality within 30 days
from the time of arrival to the emergency department, comparing ATLS
versus PTC, we will compare both the difference in mortality between
the ATLS and PTC arms as well as the difference in the change from
baseline in mortality between the ATLS and PTC arms.

For the intervention arms the change from baseline will be calculated
as the difference between the one month period of data collection
before the training was undertaken and the three month period after
the training. For the control arm the data collection period will be
four months and the difference from baseline will be calculated as the
difference between the first one month and the following three months.

>>>>

<!--
## Corresponding author
Martin Gerdin Wärnberg (MGW)

@@

-->

@@

<!--
## Word Count
4494
 -->

## Role of study sponsor and funders
The funding sources had no role in the design of this study nor during
its execution, analyses, interpretation of the data, or decision to
submit the results.

<!-- \newpage -->

@@

**Introduction** Several trauma life support programs aim to improve trauma outcomes but there is no evidence from controlled trials to show the effect of these programs on patient outcomes. We conducted a pilot study to assess the feasibility of conducting a cluster randomised controlled trial comparing the effect of Advanced Trauma Life Support (ATLS) and Primary Trauma Care (PTC) with standard care on patient outcomes. **Methods and analysis** We piloted a pragmatic three-armed parallel, cluster randomised, controlled trial in India, where neither of these programs are routinely taught. We recruited tertiary hospitals and included trauma patients and residents managing these patients. Two hospitals were randomised to ATLS, two to PTC, and three to standard care. The primary outcome was all cause mortality at 30 days from the time of arrival to the emergency department. We conducted community consultations in parallel with the pilot trial. **Ethics and dissemination** We obtained ethical approval from all participating centres. **Results** Between `r start.date` and `r end.date` we included `r n.patients` patients and `r n.residents` residents from `r n.centres` centres. The patient recruitment rate was X and the resident recruitment rate was `r recruitment.rate.residents`%. The lost to follow up rate was `r rate.lost.to.follow.up`%. `r stringr::str_to_sentence(english::english(n.m30d.atls))` (`r p.m30d.atls`%) patients died within 30 days in the ATLS arm, `r n.m30d.ptc` (`r p.m30d.ptc`%) patients in the PTC arm, and `r n.m30d.control` (`r p.m30d.control`%) patients in the standard care arm. **Conclusions** Conducting a full-scale cluster randomised controlled trial comparing the effects of ATLS, PTC, and standard care on patient outcomes should be feasible after incorporating key lessons from this pilot.

@@

<!--
# Article Summary

@@

- Cluster randomized controlled trial comparing the effect of ATLS, PTC and standard care on patient and provider outcomes.
- Prospective data collection with direct observations by dedicated project officers.
- Participating centers' heterogeneity may affect the study estimates and bias the results.
-->
\newpage

@@

We will perform a pilot study that aims to assess the feasibility of
conducting a cluster randomised controlled trial comparing ATLS and
PTC with standard care. Recent methodological guidelines indicate that
the design of efficient cluster randomised controlled trials requires
data on probable or target effect sizes, proportion of participants
with the outcome (if binary), and the intracluster correlation
coefficient [@Hemming2017]. The objectives of this pilot study were
to:

@@

## Trial Design
We piloted a three-armed cluster randomised controlled trial. There
were two intervention arms, ATLS and PTC training, and one control
arm, standard care. We collected data for four months in all three
arms, first during a one month observation phase and then during a
three month intervention phase (or continued observation in the
control arm). This design allowed us to assess outcomes both as
final values and as change from baseline.

Our study is a pilot study because its objectives involves estimating
quantities, such as the probable effect sizes, proportion of
participants with the outcome (if binary), and the intracluster
correlation coefficient, needed for the sample size calculations of a
full-scale trial [@Hemming2017]. This pilot study also establish how
many participants that can be enrolled, as well as likely drop out
rates, and the feasibility of collecting primary and secondary
outcomes.

## Study Setting
We conducted this pilot in seven Indian tertiary hospitals, where
neither ATLS, PTC, nor any other trauma life support training program
is routinely taught. The seven hospitals were King Edward Memorial
Seth G. S. Medical College and K.E.M. Hospital, Mumbai; Lokmanya Tilak
Municipal Medical College and General Hospital, Mumbai; HBT Medical
College And Dr. R N Cooper Municipal General Hospital, Mumbai; Medica
Superspecialty Hospital, Kolkata; Medical College Kolkata, Kolkata;
Sir Nil Ratan Sircar Medical College & Hospital, Kolkata; Postgraduate
Institute of Medical Education & Research, Chandigarh.

<!--
The trauma system is still developing, with
limited prehospital care, and the in hospital trauma mortality as well
as the proportion of preventable deaths remain high. Lack of standard
trauma training for healthcare providers, limited hospital resources,
inadequate processes of care, overcrowding emergency departments - are
some of the factors that contribute to the high mortality and
morbidity. During recent years efforts have been made to improve
hospital trauma care, through capacity building for trained trauma
care providers, augmenting facilities, and developing care protocols
within the hospitals. 
-->

## Eligibility Criteria for Participants and Clusters
There were two groups of participants: patients and resident doctors.

### Patient Participants
Adults (15 years or older) who present to the emergency department at
participating hospitals with a history of trauma. History of trauma is
here defined as having any of the external causes of morbidity and
mortality listed in block V01-Y36, chapter XX of the International
Classification of Disease version 10 (ICD-10) codebook as reason for
presenting. We explored intervention effects across the following
clinical subgroups: men, women, blunt multisystem trauma, penetrating
trauma, shock, severe traumatic brain injury, and elderly, as defined
by Hornor et al [@Hornor2018]. The consent form for patients are
available as Supplemental Material 1.

### Resident Doctor Participants
Resident doctors doing their speciality training in surgery or
emergency medicine managing trauma patients in the emergency
department and who were expected to remain in the participating
hospitals for at least one year. To facilitate administration each
surgical department is divided into units, which manages the out
patient department, emergency department, operating rooms etc on
different days each week. One or two, out of typically six, units'
residents were selected from each hospital. One unit consists of at
least three faculty and three to twelve residents.

To be eligible, units had to have a maximum of 25% of the doctors
trained in either ATLS, PTC, or similar training programs before the
start of the pilot. Those residents who had received training in the
last five years were considered as trained. The figure of 25% was
decided through consensus in the research team, to balance feasibility
and contamination of results.

Consent was sought from the residents in each of the intervention
groups before they underwent the ATLS or PTC training. The consent
form for residents is available as Supplemental Material 2. We did
not ask for consent from residents at the units in the control
hospitals as their practice was not affected by this pilot and we
did not collect any personal identifiable data on them.

### Clusters
Indian tertiary care hospitals that admit 400-800 adult patients with
trauma each year. We randomised on the cluster (hospital) level to
avoid contamination between intervention and control arms. To be
eligible for inclusion hospitals had to provide the following services
round the clock: operation theatres, X-ray, CT, and ultrasound
facilities, and blood bank. In addition the baseline admission rate
had to be more than 35 adult patients with major trauma per month.

## Interventions
In each intervention arm one or two units', out of typically six,
residents per hospital providing emergency care to trauma patients
were trained in either ATLS or PTC. For the purpose of this pilot
study, we will target to train a minimum of 75% of residents in each
unit. If residents dropped out or changed units after training but
before data collection was completed we planned to conduct additional
training if needed to meet the 75% criterion, but this was not
required. We did not train the units' faculty, as they are typically
not involved in the initial management of trauma patients.

The ATLS training was conducted in an ATLS certified training centre
in Mumbai, according to the standard ATLS curriculum
[@acsAtls2018]. The PTC training was conducted in New Delhi, according
to the standard PTC curriculum [@ptcfPtc2018]. These courses were
conducted over a period of 2.5 to 3 days. The residents certified
"pass" were considered as trained in respective courses.

@@

### Modifications
Both ATLS and PTC are standard training programs with fixed curricula
[@acsAtls2018; @ptcfPtc2018]. We did not modify the delivery or
content of these programs during this pilot.

### Adherence
The intervention was the training in either ATLS or PTC and resident
participants were required to adhere to, i.e. participate in, the
training, to be eligible for passing. We did not consider adherence to
training contents during care delivery as adherence to the trial
intervention.

@@

#### Baseline Training

**This section needs to be updated with the emergency medicine curriculum**

The care provided by all participating hospitals at baseline was based on
the training curriculum formulated by The National Medical Council of
India for post graduation in General Surgery
[@mciMsSurgery2019]. Regarding trauma, these guidelines state
that the student should:

@@

#### Standard of Care

**This section needs to be revised to include information from sites
where patients were not managed by surgical residents**

At most tertiary hospitals in India trauma patients present to the
emergency department where they are assessed by a doctor and referred
to the surgical bay for further management. In the surgical bay a
second or third year general surgery resident sees all the major
trauma and provide the initial care, including initiating treatment
and investigations.  This resident informs the consultant on call who
is generally an Assistant Professor. Most procedures like intercostal
drainage, open wound suturing, intubation etc. would be done in the
surgical resuscitation area, by the surgical resident.

@@

## Outcomes
Our pilot study included both participant and feasibility
outcomes. Prior to deciding on these participant outcomes we searched
the Core Outcome Measures in Effectiveness Trials (COMET) Initiative's
database but were unable to identify appropriate core outcome sets for
our populations of participants.

The primary participant outcome was all cause mortality within 30 days
from the time of arrival to the emergency department. The primary
outcome and most secondary outcome were assessed and compared both as
final values and as change from baseline. All outcomes that pertain to
the individual participant level are detailed in Supplemental Material
3. We decided to include a large number of outcomes,
including some more exploratory, so that we can test their feasibility
and relevance.

We also assessed the following feasibility outcomes, which pertain
both to overall study population as well as to the individual cluster
level:

- Recruitment rate. For both patients and residents this was equal to
  the proportion of participants enrolled, out of the total number of
  eligible participants, over the course of the pilot study.
- Lost to follow up rate. This applied only to patients and was equal
  to the proportion of patients that do not complete 30 day follow up,
  out of all enrolled patients, over the course of the pilot study.
- Pass rate. This applied only to residents in the intervention
  arms and equal the proportion of residents that pass the training
  programme, out of the total number of trained residents, over the
  course of the pilot study.
- Missing data rate. This applied to each outcome and variable and was
  equal to the proportion of missing data, over the course of the
  pilot study.
- Differences in distributions of observed and extracted data. This
  applied to each outcome and variable and compared the distributions
  of data collected by observations versus extracted from hospital
  records. For quantitative variables this was be the difference in
  means, standard deviations, medians, interquartile ranges, and
  ranges. For qualitative variables this was the differences in
  absolute counts and percentages, across categories.

@@

### Patients
Patients were screened for eligibility as they arrive at the emergency
department. Eligible patients were approached in the emergency
department to consent to follow up, if they are conscious. If they
were unconscious a patient representative was approached to consent to
follow up. Once the patient is conscious we approached the patient to
affirm the patient representative's consent. We followed up patients
at discharge, at 24 hours after arrival at the emergency department,
and at 30 days after arrival at the emergency department.

### Residents
Participating units were be screened for eligibility once hospitals
confirmed their participation. All residents in eligible units were
approached to consent to training if their hospital was randomised to
either of the intervention arms. Training was conducted approximately
one month after the study started.

## Sample size
Given budget and time constraints, including the rotation of units in
the emergency departments in Indian hospitals (which often happen on a
six months basis) the feasible data collection period was four
months. We assumed that each of the units saw 2-4 trauma patients per
week and based on that estimated that if we selected a minimum of one
unit per hospital then each hospital would enrol 8-16 patients per
month and 32-64 patients during the four months of this pilot. With a
20% attrition rate we expected each hospital to enrol 26-51 patients,
coming to a total sample size of between 156 and 306 patients for this
pilot study.

## Recruitment
To ensure adequate recruitment we only approached hospitals that
stated that the their volumes were high enough to allow us to reach
the sample size goals detailed above. Patients were enrolled by a
dedicated project officer as they arrived at the emergency
department. The recruitment period was four months. Recruitment was
monitored weekly through online conferences.  No financial or
non-financial incentives were provided to trial investigators or
participants for enrolment.

@@

### Implementation
The random allocation sequence was generated by MGW. Clusters were
enrolled by the project's core team. Patient participants were
enrolled if they presented during the project officers shift. Resident
participants were enrolled if they were in the units selected for
training. The principal investigator at each hospital selected the
units for training. For patient participants consent for follow up was
sought after randomisation from patients or patient relatives as
appropriate. For resident participants consent was also sought after
randomisation.

## Blinding
It was not possible to blind investigators or participants to
interventions. We did not blind the data analysts during this pilot,
but we plan to blind the data analysts during the full-scale trial.

## Data Collection
Data collection started one month before the training was delivered,
to establish a baseline. Each participating hospital had a dedicated
project officer who collected the data.

Because participating residents were assigned designated days for
trauma care for a period of 6 months, data were collected during those
particular days and shifts when these trained doctors were in the
emergency department. The project officers collected data both by
observing the care delivered and by interviewing the participants, and
by extracting data from hospital records.

Data collection continued for three months after training. The
research officers collected data of all trauma patients who presented
to the participating units during their duty hours. Those patients who
were admitted were followed up for complications and other in-hospital
outcome measures, for example length of stay. Patients who were not
admitted were followed up telephonically for mortality outcomes and
quality of life outcomes. The follow up period was be 30 days. The
project officers made at most three attempts to reach a participant or
participant representative telephonically, after which the data were
recorded as missing.

The project officer administered the study information and informed
consent (consent was sought only for data collection including follow
up) to the patient, or the patient's representative as appropriate,
once the patient was stabilised. They will continue to collect data
once they have received the consent.

Details of data of those patients/relatives not willing to give
consent were not included in the analysis. The number of patients who
opt out from data collection was collected.

## Variables
The project officers collected data on demographics, time of injury to
arrival at the participating hospital, time to recording vital signs,
vital signs, and times to and management details including imaging and
surgery. Details of any injury sustained were collected and coded
using ICD 10 and the Abbreviated Injury Scale (AIS). For ICD 10 coders
underwent the WHO online ICD 10 training module and for AIS they were
accredited. Based on AIS we calculated the Injury Severity Score
(ISS) and the New ISS (NISS). Supplemental Material 4 contains a full
variable list, with definitions.

## Patient and public involvement
In this study, we conducted community consultations to collect
inputs from patients, their caregivers, patient groups, and resident
doctors to be used in the selection of outcome measures and
implementation of the full-scale trial; following the Guidance for
Reporting Involvement of Patients and the Public (GRIIPS 2)
[@Staniszewska2017].

During the pilot study, interviews were conducted with post-discharge
trauma patients and their caregivers to identify the outcomes most
relevant to them. These patients were identified through the medical
registers of the participating hospitals, contacted through telephone,
and after receiving their consent they were interviewed as per their
convenience. Their consent form is available as Supplemental
Material 5.  Additionally, members from non-government organizations
working with trauma patients and the hospital Social Service Section
were contacted for their views on contextual patient-centred outcomes
for trauma patients. Their consent form is available as Supplemental
Material 6. For feasibility, these interviews were held in each of the
cities where the participating centres are located. The commonest
patient-centred outcomes reported across all the locations will be
incorporated into the evaluation of the effects of the different
training programs and standardized care on patient outcomes.

Similarly, the inputs of resident doctor participants at each
participating centre were collected during the pilot study. A
discussion with the residents were conducted to document any
challenges or suggestions they may have in the scheduling or
implementation of the training programs. These inputs will be
incorporated in the final study.

A summary of the findings of the study as well as their inputs will be
shared with those who participated in the interviews and
discussions. A meeting will be held with the patient participants, at
each city, where the changes in the measured patient-centered outcomes
would be presented to them. Another meeting will be held with the
resident doctors at each hospital to present the confidence of the
residents after being trained.  Any suggestions and reflections from
the participants during the meetings will be used as inputs for
planning the final study.

## Data management
We used the online data collection tool KoboToolbox [@KoboToolbox],
accessible only over a virtual private network (VPN), for each
participating hospital to upload pseudonymised data to secure
servers. Data validation techniques such as restricted values or
values within a specific range were used to avoid ambiguous data
entries and ensure the validity of the data. Ambiguous responses and
errors were resolved after discussion with the core team during weekly
meetings. An instruction manual or codebook for data variables was
prepared to ensure consistency in data entry. This manual was referred
to during the project data collection and variable descriptions were
visible for each variable in the online data collection
tool. Pseudonymised data were stored at the centralised server. The
data were only accessible by the project's principal investigator or
by delegation by the project principal investigator only.

## Data monitoring
Weekly meetings with the core team and project officers took place. We
conducted an interim analysis on October 12, 2022, and decided to
complete the study as participants were consenting to be included in
the study and key variables including mortality outcomes could be
collected. No outcomes were dropped. We did not use a data monitoring
committee in the pilot study due to its limited scope.

## Statistical Methods
We analysed all pilot data using descriptive statistics. Quantitative
variables are summarised as mean +/- standard deviation, median,
interquartile range and range. Qualitative variables are presented as
absolute numbers and percentages. Feasibility outcomes are summarised
both on the overall sample level as well as on the individual cluster
level. We used an empty generalised linear mixed model<!-- see
[@Sommet2017] --> to estimate the intracluster correlation
coefficient.

We compared participant outcomes in three combinations of trial arms:
ATLS versus PTC, ATLS versus standard care, and PTC versus standard
care. In each combination we compared both differences in final values
and differences in change from baseline. For the intervention arms the
change from baseline was calculated as the difference between the one
month period of data collection before the training was undertaken and
the three month period after the training. For the control arm the
data collection period was four months and the difference from
baseline was calculated as the difference between the first one
month and the following three months.

Within each combination of trial arms we conducted subgroup analyses
of men, women, blunt multisystem trauma, penetrating trauma, shock,
severe traumatic brain injury, and elderly. Table S7.1 in Supplemental
Material 7 shows which outcomes that were assessed in what subgroups,
decided through consensus in the research team. We further compared
the results of all subgroups with the results in the whole cohort, and
compared the results in the female subgroup with the male subgroup,
and the results in the blunt multisystem trauma subgroup with the
penetrating trauma subgroup. We are aware that the numbers in some of
these subgroups are small, but we included them to help guide the
formulation of the statistical analysis plan for the full-scale trial.

We calculated both absolute and relative differences for each
comparison, along with 75, 85, and 95% confidence intervals. We used
an empirical bootstrap procedure with 1000 draws to estimate these
confidence intervals. We did not perform any formal hypothesis tests
during the analysis of this pilot's data [@Lee2014]. We also compared
the data collected through observations and interviews with the data
collected from hospital records, to assess the feasibility of
collecting data from hospital records in the full-scale trial.
